---
title: "Hands-on Session I: Proteomic and Metabolomic Characterization of COVID-19 Patient Sera"
author: "Domenica Arlet Teller Guardián (domenicaarlet.teller01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    fig_caption: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6)
```

# Introduction

This practical focuses on data wrangling, exploratory data analysis, visualization, and critical comparison with published results using real biomedical data.

We will work with the publication [**Proteomic and Metabolomic Characterization of COVID-19 Patient Sera**](https://www.sciencedirect.com/science/article/pii/S0092867420306279)

# Exercise 1: Data Loading and Cleaning

**Download and load Supplementary Table 1 (“Additional Demographical and Baseline Characteristics of COVID-19 Patients and Control Groups”) from the supplementary materials of the publication.**

The study analyzed proteomic and metabolomic profiles of sera from 46 COVID-19 patients and 53 control individuals. Using machine learning, the authors identified 22 proteins and 7 metabolites that could classify severe versus non-severe patients. The data used here comes from Supplementary Table S1, which contains demographic and clinical laboratory information for 118 individuals across four groups: healthy controls, non-COVID-19 patients, non-severe COVID-19, and severe COVID-19. The dataset includes 31 variables covering patient identifiers, dates, demographic characteristics, and 12 clinical laboratory measurements.

```{r load-libraries}
library(tidyverse)
library(readxl)
library(janitor)
library(naniar)
library(rstatix)
library(ggpubr)
library(pheatmap)
library(table1)
library(htmltools)
library(xml2)
library(rvest)
```

```{r load-data}
data <- read_excel("data/raw_data/Table_S1.xlsx", sheet = "Clinical_information")

dim(data)
str(data)
```

The dataset contains `r nrow(data)` samples and `r ncol(data)` variables.

```{r clean-dates}
# Date cleaning & format harmonization

# Fix dates stored as Excel serial numbers
data$`Date of progression to severe state` <- as.Date(
  as.numeric(data$`Date of progression to severe state`),
  origin = "1899-12-30"
)

data$`Sampling date for proteomics` <- as.Date(
  as.numeric(data$`Sampling date for proteomics`),
  origin = "1899-12-30"
)

# Replace "/" entries with NA in character columns
cols_with_slash <- names(which(
  sapply(data, function(col) is.character(col) && any(col == "/", na.rm = TRUE))
))

data[cols_with_slash] <- lapply(data[cols_with_slash], function(col) {
  col[col == "/"] <- NA
  col
})

# Convert POSIXct columns to Date
posix_cols <- sapply(data, inherits, what = "POSIXt")
data[posix_cols] <- lapply(data[posix_cols], as.Date)
```

```{r prepare-eda-dataset}
# Save the full dataset
data_full <- data

# Remove technical ID variables not useful for EDA
data_eda <- data_full %>%
  select(-`MS ID b`, -`MSRep ID c`, -`Metabolomics ID e`)

# IMPORTANT: Clean column names FIRST, THEN encode factors
# (doing it the other way round causes factor levels to become NA)
data_eda <- data_eda %>%
  clean_names(replace = c("μ" = "u"))

# Now encode factors using the CLEANED column names
data_eda$group_d <- factor(
  data_eda$group_d,
  levels = c(0, 1, 2, 3),
  labels = c("healthy", "non-COVID-19", "non-Severe", "Severe")
)

data_eda$sex_g <- factor(
  data_eda$sex_g,
  levels = c(0, 1),
  labels = c("Female", "Male")
)

# Ensure numeric types for BMI and CRP (may have been read as character)
data_eda$bmi_h <- as.numeric(data_eda$bmi_h)
data_eda$crp_i_mg_l <- as.numeric(data_eda$crp_i_mg_l)

# Drop unused factor levels
data_eda$group_d <- droplevels(data_eda$group_d)

# Verify the factor encoding worked
table(data_eda$group_d, useNA = "always")
table(data_eda$sex_g, useNA = "always")

# Define the 12 clinical variables for downstream analyses
clinical_variables <- c(
  "bmi_h",
  "wbc_count_109_l",
  "lymphocyte_count_109_l",
  "monocyte_count_109_l",
  "platelet_count_109_l",
  "crp_i_mg_l",
  "alt_j_u_l",
  "ast_k_u_l",
  "ggt_l_u_l",
  "tbil_m_umol_l",
  "dbil_n_umol_l",
  "creatinine_umol_l",
  "glucose_mmol_l"
)

# Readable labels for clinical variables (used in plots)
clinical_labels <- c(
  "wbc_count_109_l"        = "White blood cell (WBC)",
  "lymphocyte_count_109_l" = "Lymphocyte",
  "monocyte_count_109_l"   = "Monocyte",
  "platelet_count_109_l"   = "Platelet",
  "crp_i_mg_l"             = "C-reactive protein (CRP)",
  "alt_j_u_l"              = "Alanine aminotransferase (ALT)",
  "ast_k_u_l"              = "Aspartate aminotransferase (AST)",
  "ggt_l_u_l"              = "Glutamyltransferase (GGT)",
  "tbil_m_umol_l"          = "Total bilirubin (TBIL)",
  "dbil_n_umol_l"          = "Direct bilirubin (DBIL)",
  "creatinine_umol_l"      = "Creatinine",
  "glucose_mmol_l"         = "Glucose"
)

colnames(data_eda)
```

------------------------------------------------------------------------

## Exercise 1.1

Reproduce Table 1 from the manuscript using the supplementary data. Match the reported summary statistics as closely as possible.

Briefly discuss any discrepancies between your results and the published table, and provide possible explanations.

```{r table1-render}
# Custom render functions to match the paper's Table 1 format

my.render.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x)), {
    out <- c("",
             "Mean ± SD"    = sprintf("%s ± %s", MEAN, SD),
             "Median (IQR)" = sprintf("%s (%s–%s)", MEDIAN, Q1, Q3),
             "Range"        = sprintf("%s–%s", MIN, MAX))
    out[names(out) != "Missing"]
  })
}

my.render.cat <- function(x) {
  c("", sapply(levels(x), function(lv) {
    n <- sum(x == lv, na.rm = TRUE)
    N <- sum(!is.na(x))
    sprintf("%d (%.1f)", n, 100 * n / N)
  }, USE.NAMES = TRUE))
}
```

```{r table1-prepare}
# Create a duplicate set of COVID-19 rows to form the "Total COVID" column

data_tab1 <- bind_rows(
  data_eda %>%
    mutate(group_tab1 = as.character(group_d)),

  data_eda %>%
    filter(group_d %in% c("non-Severe", "Severe")) %>%
    mutate(group_tab1 = "Total COVID")
)

data_tab1$group_tab1 <- factor(
  data_tab1$group_tab1,
  levels = c("healthy", "non-COVID-19", "Total COVID", "non-Severe", "Severe")
)

# Compute time-to-event variables
# The paper reports integer day counts; we compute them from dates.
# A +1 adjustment is applied to match the convention where same-day = 1 day.

data_tab1 <- data_tab1 %>%
  mutate(
    time_from_onset_to_admission_days =
      ifelse(
        !is.na(onset_date_f) & !is.na(admission_date),
        as.numeric(admission_date - onset_date_f) + 1,
        NA_real_
      ),

    time_from_admission_to_severe_days =
      ifelse(
        group_d == "Severe" &
          !is.na(admission_date) &
          !is.na(date_of_progression_to_severe_state),
        as.numeric(date_of_progression_to_severe_state - admission_date) + 1,
        NA_real_
      )
  )
```

```{r table1-generate}
# Build Table 1 with available variables

table_1_data <- data_tab1 %>%
  select(
    group_tab1,
    sex_g,
    age_year,
    bmi_h,
    time_from_onset_to_admission_days,
    time_from_admission_to_severe_days
  )

# Label variables for display
label(table_1_data$sex_g) <- "Sex - no. (%)"
label(table_1_data$age_year) <- "Age - year"
label(table_1_data$bmi_h) <- "BMI, kg/m²"
label(table_1_data$time_from_onset_to_admission_days) <- "Time from Onset to Admission, Days"
label(table_1_data$time_from_admission_to_severe_days) <- "Time from Admission to Severe, Days"

table1(
  ~ sex_g +
    age_year +
    bmi_h +
    time_from_onset_to_admission_days +
    time_from_admission_to_severe_days |
    group_tab1,
  data = table_1_data,
  render.continuous = my.render.cont,
  render.categorical = my.render.cat,
  overall = FALSE,
  topclass = "Rtable1-zebra",
  caption = "Table 1. Demographic and Baseline Characteristics by Disease Group"
)
```

### Discrepancies with the published Table 1

The supplementary data (Table S1) only provides demographic variables (sex, age, BMI) and clinical laboratory measurements. It does not include smoking status, alcohol use, symptoms (fever, cough, headache, fatigue, pharyngalgia, expectoration, diarrhea, chest tightness), chest CT findings, comorbidities (hypertension, diabetes, respiratory, endocrine, kidney, digestive), oxygenation index, or treatment variables (oxygen inhalation, antibiotics, antiviral drugs, immunoglobulin, methylprednisolone, Chinese medicine). Therefore, those rows of the published Table 1 cannot be reproduced from the available supplementary data alone.

For the variables that can be compared, the reproduced statistics closely match the published values. Minor numerical discrepancies may arise from differences in rounding conventions and the handling of time-to-event calculations. The paper reports "Time from Onset to Admission" with a mean of 6.1 ± 4.4 days for total COVID-19 patients; depending on whether the calculation includes or excludes the onset/admission day itself (±1 day), slight differences can appear. Overall, the reproducible statistics (sex distribution, age, BMI) are consistent with the published table.

------------------------------------------------------------------------

## Exercise 1.2

Perform an exploratory data analysis (EDA) of the dataset. At a minimum, include:

-   Descriptive statistics for key variables\
-   Appropriate visualizations (e.g. distributions, group comparisons)\
-   A brief written summary of notable patterns, anomalies, or data quality issues

### Missing data analysis

Missing data analysis is performed to quantify overall missingness, assess whether missingness differs by patient group, and determine whether missing values are structurally driven (due to clinical relevance) rather than data quality issues.

```{r missing-data}
# Missing values per variable
missing_summary <- data_eda %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_n") %>%
  mutate(missing_pct = round(missing_n / nrow(data_eda) * 100, 1))

missing_summary %>% filter(missing_n > 0)

# Lollipop plot of missing data
gg_miss_var(data_eda)
```

Date of progression to severe state has the highest missingness (\>75%), followed by sampling dates, cohort assignment, onset date and discharge date. These are all variables that apply only to specific patient subgroups.

```{r missing-by-group}
# Missing data by patient group

missing_by_group <- data_eda %>%
  group_by(group_d) %>%
  summarise(
    across(everything(), ~ round(mean(is.na(.)) * 100, 1), .names = "{.col}")
  )

missing_long <- missing_by_group %>%
  pivot_longer(cols = -group_d, names_to = "variable", values_to = "missing_pct")

ggplot(missing_long, aes(x = variable, y = group_d, fill = missing_pct)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Missing Data Heatmap by Patient Group",
    x = "Variable", y = "Patient Group", fill = "% Missing"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The heatmap confirms that variables with substantial missingness are primarily associated with clinical events not applicable to healthy or non-COVID-19 patients. The healthy group has 100% missing data in almost every clinical variable except sex and age. This pattern shows that missing values are structurally induced rather than arising from data collection errors.

### Descriptive statistics

```{r descriptive-stats}
# Group-wise descriptive statistics using median and IQR (robust to skewness)

group_clinical_summary <- data_eda %>%
  group_by(group_d) %>%
  summarise(
    across(
      all_of(clinical_variables),
      list(
        median = ~ median(.x, na.rm = TRUE),
        q1     = ~ quantile(.x, 0.25, na.rm = TRUE),
        q3     = ~ quantile(.x, 0.75, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

group_clinical_summary
```

### Normality assessment

Distributional properties of clinical variables are assessed using histograms, Q-Q plots, and Shapiro-Wilk tests to determine whether parametric or non-parametric methods are appropriate.

```{r long-clinical}
# Reshape clinical variables to long format for unified plotting

long_clinical <- data_eda %>%
  pivot_longer(
    cols = all_of(clinical_variables),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value))
```

```{r histograms, fig.width=10, fig.height=8}
ggplot(long_clinical, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "grey80", color = "black") +
  geom_density(color = "red", linewidth = 1) +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distribution of Clinical Variables", x = "Value", y = "Density") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Most variables show right-skewed distributions (ALT, AST, CRP, GGT, bilirubin, glucose), while BMI and platelet count appear approximately symmetric.

```{r qq-plots, fig.width=10, fig.height=8}
ggplot(long_clinical, aes(sample = value)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Q–Q Plots of Clinical Variables") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

```{r shapiro}
normality_tests <- long_clinical %>%
  group_by(variable) %>%
  summarise(
    n = n(),
    shapiro_p = ifelse(n <= 5000, shapiro.test(value)$p.value, NA_real_)
  )

normality_tests
```

The Shapiro-Wilk test confirms that most clinical variables deviate significantly from normality (p \< 0.05), with the exception of BMI and platelet count. Given these results, non-parametric methods will be used for group comparisons.

### Group comparisons

```{r boxplots-all, fig.width=10, fig.height=8}
ggplot(long_clinical, aes(x = group_d, y = value, color = group_d)) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Distribution of Clinical Variables by Patient Group",
    x = "Patient Group", y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

Clear severity-dependent trends are observed: CRP, creatinine, and glucose increase from healthy controls to severe COVID-19, while lymphocyte and platelet counts decrease in severe cases.

```{r log-boxplots, fig.width=10, fig.height=6}
# Log-scaled boxplots for the most skewed variables

skewed_vars <- c("crp_i_mg_l", "alt_j_u_l", "ast_k_u_l",
                 "ggt_l_u_l", "tbil_m_umol_l", "dbil_n_umol_l")

ggplot(
  filter(long_clinical, variable %in% skewed_vars),
  aes(x = group_d, y = value, fill = group_d)
) +
  geom_boxplot(outlier.alpha = 0.4) +
  scale_y_log10() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Log-Scaled Distribution of Skewed Clinical Variables",
    x = "Patient Group", y = "Value (log scale)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

After log transformation, group differences in CRP, ALT, AST, and bilirubin become more apparent, with severe patients showing consistently elevated values.

```{r individual-plots, fig.width=10, fig.height=4}
# Individual key variable plots

p1 <- ggplot(
  filter(long_clinical, variable == "crp_i_mg_l"),
  aes(x = group_d, y = value, fill = group_d)
) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  scale_y_log10() +
  labs(title = "CRP", x = "", y = "CRP (mg/L, log scale)") +
  theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(
  filter(long_clinical, variable == "lymphocyte_count_109_l"),
  aes(x = group_d, y = value, color = group_d)
) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  labs(title = "Lymphocyte Count", x = "", y = "×10⁹/L") +
  theme_minimal() + theme(legend.position = "none")

p3 <- ggplot(
  filter(long_clinical, variable == "platelet_count_109_l"),
  aes(x = group_d, y = value)
) +
  geom_boxplot(fill = "grey80") +
  stat_summary(fun = median, geom = "text",
               aes(label = round(after_stat(y), 1)),
               vjust = -0.6, size = 3) +
  labs(title = "Platelet Count", x = "", y = "×10⁹/L") +
  theme_minimal()

ggarrange(p1, p2, p3, ncol = 3)
```

CRP is markedly elevated in severe COVID-19 patients; lymphocyte counts show progressive severity-dependent decline (lymphopenia); platelet counts are reduced in severe cases (thrombocytopenia).

### Statistical tests

```{r kruskal-wallis}
# Kruskal-Wallis test: is at least one group different?

kruskal_results <- long_clinical %>%
  group_by(variable) %>%
  kruskal_test(value ~ group_d)

kruskal_results
```

```{r dunn-test}
# Dunn post-hoc test with BH correction: which pairs differ?

dunn_results <- long_clinical %>%
  group_by(variable) %>%
  dunn_test(value ~ group_d, p.adjust.method = "BH")

sig_dunn <- dunn_results %>% filter(p.adj < 0.05)
sig_dunn
```

Kruskal-Wallis tests revealed significant differences across disease groups for CRP, glucose, lymphocyte count, and white blood cell count (p \< 0.05). Post-hoc Dunn tests confirmed strong pairwise differences for CRP, glucose, and lymphocyte count, with the strongest contrasts between severe patients and both non-COVID-19 and non-severe groups.

### Sex as a potential confounder

```{r sex-analysis, fig.width=10, fig.height=8}
ggplot(long_clinical, aes(x = sex_g, y = value, fill = sex_g)) +
  geom_boxplot(outlier.alpha = 0.4, alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Distribution of Clinical Variables by Sex",
    x = "Sex", y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

```{r sex-wilcox}
sex_comparison <- long_clinical %>%
  group_by(variable) %>%
  wilcox_test(value ~ sex_g) %>%
  adjust_pvalue(method = "BH") %>%
  select(variable, n1, n2, p, p.adj)

sex_comparison
```

Most clinical variables show substantial overlap between sexes. Significant differences were observed for ALT, AST, creatinine, and bilirubin, consistent with known physiological sex differences. Overall, sex is unlikely to act as a major confounder in severity-based analyses.

### EDA Summary

Data cleaning ensured consistent variable formats, correct date handling, and appropriate factor definitions. Variables with high missingness were structurally driven (not applicable to healthy or non-severe patients) rather than data quality issues.

Most clinical variables showed strong right-skewness, confirmed by histograms, Q-Q plots, and Shapiro-Wilk tests. Non-parametric methods were therefore used for group comparisons.

Comparative analyses revealed significant differences across disease groups for CRP, glucose, lymphocyte count, AST, and WBC count, with the strongest contrasts between non-severe and severe COVID-19 patients. Sex-based analyses showed limited effects overall, with significant differences only in a subset of laboratory parameters consistent with known physiological sex differences.

------------------------------------------------------------------------

# Exercise 2

## Exercise 2.1

**Reproduce Supplementary Figure 1 using the data without modifying or removing any observations. Use the same variables and groupings as in the original figure.**

The original Figure S1 shows boxplots with jittered points for 12 clinical variables across three groups (non-COVID-19, non-Severe, Severe), with significance brackets from unpaired Welch's t-tests.

```{r fig-s1-raw, fig.width=12, fig.height=10}
# Filter to the three groups in Figure S1
df_s1 <- data_eda %>%
  filter(group_d %in% c("non-COVID-19", "non-Severe", "Severe")) %>%
  droplevels()

# Pivot to long format with readable labels
df_long_s1 <- df_s1 %>%
  pivot_longer(
    cols = all_of(names(clinical_labels)),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  mutate(variable = factor(variable,
                           levels = names(clinical_labels),
                           labels = clinical_labels))

# Define pairwise comparisons matching the original figure
my_comparisons <- list(
  c("non-COVID-19", "non-Severe"),
  c("non-COVID-19", "Severe"),
  c("non-Severe", "Severe")
)

# Create individual plots
plot_list <- lapply(levels(df_long_s1$variable), function(var) {
  d <- filter(df_long_s1, variable == var)
  ggboxplot(d, x = "group_d", y = "value",
            color = "group_d", add = "jitter",
            add.params = list(size = 1.5, alpha = 0.5)) +
    stat_compare_means(comparisons = my_comparisons,
                       method = "t.test",
                       label = "p.signif",
                       size = 3) +
    labs(title = var, x = "", y = "") +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          plot.title = element_text(size = 9, face = "bold"))
})

# Arrange in a 4x3 grid (matching the original figure layout)
ggarrange(plotlist = plot_list, ncol = 4, nrow = 3)
```

The reproduced figure shows the same patterns as the original: significant lymphopenia and suppressed monocyte counts in severe patients, elevated CRP and AST, and increased glucose levels. The significance brackets match the asterisk annotations in the published Figure S1.

------------------------------------------------------------------------

## Exercise 2.2

**Identify and handle outliers using an appropriate and well-justified method (e.g. removal, transformation, or robust statistics).**

**Reproduce Supplementary Figure 1 after outlier handling and discuss how and why the results change.**

We use **winsorization** (capping extreme values at 1.5×IQR boundaries) rather than removal, because clinical outliers in biomedical data often represent real pathological states rather than measurement errors. Removing them could bias results toward milder presentations.

```{r outlier-identification}
# Count outliers per variable using the IQR criterion

outlier_counts <- df_s1 %>%
  pivot_longer(cols = all_of(names(clinical_labels)),
               names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(variable) %>%
  summarise(
    n = n(),
    n_outliers = {
      q <- quantile(value, c(0.25, 0.75))
      iqr_val <- q[2] - q[1]
      sum(value < q[1] - 1.5 * iqr_val | value > q[2] + 1.5 * iqr_val)
    },
    pct_outliers = round(n_outliers / n * 100, 1),
    .groups = "drop"
  )

outlier_counts
```

```{r winsorize-and-replot, fig.width=12, fig.height=10}
# Winsorize function: cap values at Q1 - 1.5*IQR and Q3 + 1.5*IQR
winsorize_iqr <- function(x) {
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
  iqr_val <- q[2] - q[1]
  lower <- q[1] - 1.5 * iqr_val
  upper <- q[2] + 1.5 * iqr_val
  x[x < lower & !is.na(x)] <- lower
  x[x > upper & !is.na(x)] <- upper
  x
}

# Apply winsorization
df_s1_win <- df_s1 %>%
  mutate(across(all_of(names(clinical_labels)), winsorize_iqr))

# Pivot winsorized data to long format
df_long_win <- df_s1_win %>%
  pivot_longer(
    cols = all_of(names(clinical_labels)),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  mutate(variable = factor(variable,
                           levels = names(clinical_labels),
                           labels = clinical_labels))

# Reproduce Figure S1 after winsorization
plot_list_win <- lapply(levels(df_long_win$variable), function(var) {
  d <- filter(df_long_win, variable == var)
  ggboxplot(d, x = "group_d", y = "value",
            color = "group_d", add = "jitter",
            add.params = list(size = 1.5, alpha = 0.5)) +
    stat_compare_means(comparisons = my_comparisons,
                       method = "t.test",
                       label = "p.signif",
                       size = 3) +
    labs(title = var, x = "", y = "") +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          plot.title = element_text(size = 9, face = "bold"))
})

ggarrange(plotlist = plot_list_win, ncol = 4, nrow = 3)
```

### Discussion: Effect of outlier handling

After winsorization, the boxplots show compressed whiskers and fewer extreme points, but the overall group differences remain consistent. Variables such as CRP, glucose, and lymphocyte count retain their significant severity-dependent patterns. This confirms that the observed biological trends are robust and not driven by a few extreme observations. For variables like ALT, AST, and GGT, where highly skewed distributions produced many IQR-based "outliers," winsorization helps reveal the underlying group differences more clearly without discarding clinically meaningful data points.

------------------------------------------------------------------------

# Exercise 3

Create a heatmap of the biomarker data. Include group and gender as annotation variables.

```{r heatmap, fig.width=10, fig.height=6}
# Prepare the matrix: samples as rows, variables as columns
mat <- data_eda %>%
  select(all_of(clinical_variables)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Remove variables with >40% missing
missing_prop <- colMeans(is.na(mat))
mat <- mat[, missing_prop < 0.40]

# Impute remaining NAs with column medians
for (j in seq_len(ncol(mat))) {
  med <- median(mat[, j], na.rm = TRUE)
  mat[is.na(mat[, j]), j] <- med
}

# Z-score scale
mat_scaled <- scale(mat)

# pheatmap expects variables as rows, samples as columns
mat_t <- t(mat_scaled)

# Annotation for columns (each column = one sample)
annot <- data.frame(
  Group = data_eda$group_d,
  Sex   = data_eda$sex_g,
  row.names = paste0("S", seq_len(nrow(data_eda)))
)
colnames(mat_t) <- rownames(annot)

ann_colors <- list(
  Group = c(
    "healthy"      = "#4DAF4A",
    "non-COVID-19" = "#377EB8",
    "non-Severe"   = "#984EA3",
    "Severe"       = "#E41A1C"
  ),
  Sex = c(
    "Female" = "#FF7F00",
    "Male"   = "#A65628"
  )
)

pheatmap(
  mat_t,
  annotation_col = annot,
  annotation_colors = ann_colors,
  clustering_method = "complete",
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  show_colnames = FALSE,
  fontsize_row = 8,
  main = "Heatmap of Clinical Biomarkers by Patient Group and Sex"
)
```

The heatmap reveals that samples partially cluster by disease group, with severe COVID-19 patients showing a distinct biomarker profile characterized by elevated inflammatory markers (CRP, AST) and reduced lymphocyte and platelet counts. Healthy controls form a relatively homogeneous cluster with low values across most biomarkers. The non-COVID-19 and non-severe groups show intermediate and more heterogeneous patterns. Sex does not appear to drive the main clustering structure.

# Conclusion

**Write a concise conclusion (1 paragraphs) summarizing the main differences observed across the three patient groups based on the twelve clinical parameters. Support your statements with evidence from your analyses:**

Exploratory and comparative analyses of twelve clinical parameters revealed distinct molecular profiles across the three patient groups. Severe COVID-19 patients exhibited significantly elevated inflammatory markers (CRP, AST) and metabolic indicators (glucose), alongside suppressed lymphocyte and platelet counts, consistent with the immune dysregulation and systemic inflammation reported by Shen et al. (2020). Non-severe COVID-19 patients showed intermediate profiles, with laboratory values generally closer to the non-COVID-19 group. The non-COVID-19 control group displayed heterogeneous patterns due to diverse underlying conditions but lacked the characteristic lymphopenia and acute inflammatory signature of severe COVID-19. These findings were robust to outlier handling (winsorization) and were confirmed by both non-parametric statistical tests (Kruskal-Wallis and Dunn post-hoc) and heatmap clustering, which visually separated severe patients from other groups based on clinical biomarker profiles.

------------------------------------------------------------------------

# Session Info {.unnumbered}

```{r session-info, results='asis', echo=FALSE}
sessionInfo()
```
